---
title: "Labor and Repro Justice Survey — Data Cleaning & Quality Flagging"
output: html_document
---

## Overview

This document cleans and prepares the Labor and Repro Justice Survey data exported from Qualtrics. It proceeds in seven steps:

1. Load the required R libraries
2. Load the raw CSV and remove Qualtrics metadata rows
3. Split multi-select question columns into individual binary columns
4. Assign a response quality flag (E / P / Q) to each row
5. Add suspicion detail flags to help prioritize review of questionable responses
6. Print a summary of the flagged dataset
7. Save the cleaned, flagged dataset as a new CSV

Each section can be run independently as a chunk. Run them in order top to bottom on first use.

---

## Step 1: Load Libraries

The script uses `dplyr` for data manipulation and `stringr` for text matching. If either package is not yet installed, uncomment the install line below.

```{r setup, message=FALSE, warning=FALSE}
# install.packages(c("dplyr", "stringr"))
library(dplyr)
library(stringr)
```

---

## Step 2: Load the Raw Data

Qualtrics CSV exports always include three header rows before the actual survey responses begin:

- **Row 1** — variable names (e.g., `Q1`, `Q23`). `read.csv()` uses this as column headers.
- **Row 2** — full question text labels (e.g., "What best describes your role?"). Not needed for analysis; removed below.
- **Row 3** — Qualtrics internal import IDs (e.g., `{"ImportId":"QID2"}`). Not needed; removed below.

Actual survey responses begin on row 4 of the file.

```{r load-data}

file_path <- "C:/Users/sandh/Downloads/labor-repro-survey.csv"

df <- read.csv(file_path, header = TRUE, stringsAsFactors = FALSE, check.names = TRUE)
df <- df[-c(1, 2), ]   # remove question-label row and import-ID row
rownames(df) <- NULL   # reset row index to 1..n

# Rename the duration column (R converts spaces/parentheses to dots)
names(df)[names(df) == "Duration..in.seconds."] <- "duration_sec"
df$duration_sec <- as.numeric(df$duration_sec)

n_original_cols <- ncol(df)   # used later to confirm dummy expansion
```

```{r verify-load}
# Confirm expected dimensions before proceeding
cat("Rows (responses):", nrow(df), "\n")
cat("Columns:", ncol(df), "\n")
cat("Column names:\n")
print(names(df))
```

---

## Step 3: Split Multi-Select Columns into Binary Dummy Columns

Several survey questions allowed respondents to select more than one answer. Qualtrics stores all selected options in a single cell, separated by commas. For example, a response to the work-type question might appear as:

> `Labor organizing,Reproductive justice,Other (please specify)`

This format is not usable for analysis. The code below converts each multi-select column into a set of binary (0/1) columns — one column per response option — where 1 means the option was selected and 0 means it was not.

**How options are separated.** Qualtrics puts a comma with *no space after it* between selected options. Some option labels contain their own commas (for example, several Q15 options), but those are followed by a space (`", "`). The code therefore splits only on commas that are not followed by a space, so labels with internal commas stay intact.

**Which columns are split.** Rather than typing out every option by hand, the code finds the multi-select columns automatically: any question column (names starting with `Q` or `QID`) that contains at least one Qualtrics separator comma. Open-text columns (`*_TEXT`, Q20, Q21a, Q22) are never split. The list of columns and options it found is printed below so you can check it. If a column is wrongly included or missed, edit `multi_select_cols` by hand.

The helper function `make_dummies()` does the conversion. It takes the source column, a list of all possible options, and a prefix for the new column names. A cell gets a 1 only when one of its split pieces matches the option exactly, so an option like "Other" doesn't match "Other (please specify)".

**New column naming convention:** `<prefix>__<option_text>` with punctuation replaced by underscores.

```{r dummy-function}
option_sep <- ",(?!\\s)"   # a comma not followed by whitespace

split_options <- function(x) {
  lapply(str_split(x, option_sep), function(p) trimws(p[!is.na(p) & trimws(p) != ""]))
}

make_dummies <- function(df, source_col, options, prefix) {
  pieces <- split_options(df[[source_col]])
  for (opt in options) {
    col_name <- paste0(prefix, "__",
                       str_replace_all(trimws(opt), "[^A-Za-z0-9]+", "_"))
    col_name <- str_replace_all(col_name, "_+", "_")
    col_name <- str_remove(col_name, "_$")
    df[[col_name]] <- as.integer(sapply(pieces, function(p) opt %in% p))
  }
  df
}
```

```{r make-dummies}
open_text_cols <- c("Q20", "Q21a", "Q22")

candidate_cols <- names(df)[str_detect(names(df), "^(Q|QID)\\d") &
                            !str_detect(names(df), "_TEXT$") &
                            !names(df) %in% open_text_cols]

multi_select_cols <- candidate_cols[sapply(candidate_cols, function(col) {
  any(str_detect(df[[col]], option_sep), na.rm = TRUE)
})]

for (col in multi_select_cols) {
  options <- sort(unique(unlist(split_options(df[[col]]))))
  cat(sprintf("\n%s (%d options):\n", col, length(options)))
  cat(paste0("  - ", options, collapse = "\n"), "\n")
  df <- make_dummies(df, col, options, prefix = col)
}
```

```{r verify-dummies}
# Confirm total column count after all dummy expansion
cat("Multi-select columns split:", paste(multi_select_cols, collapse = ", "), "\n")
cat("Started with", n_original_cols, "original columns\n")
cat("Total columns after dummy expansion:", ncol(df), "\n")
```

---

## Step 4: Assign Response Quality Flags

Each response receives one of three quality flags in a new column called `response_flag`. The flags are assigned in priority order: E is checked first, then P, and Q is the default for everything remaining.

**E — Definitely Exclude**
Applied when any one of the following is true:

- `Status == "Survey Preview"`: Test submissions recorded during survey setup; not real respondents.
- `Q1 is blank`: The respondent never answered the first substantive question, indicating they did not engage with the survey content.
- `Fewer than 18 of 27 questions answered`: The survey contained 18 required questions. Any response with fewer than 18 questions answered bypassed required fields, indicating automated or invalid form submission rather than genuine participation.
- `Contradictory volunteer/staff answers`: The respondent answered Q1 with "Member of an all-volunteer network" but checked "Staff" in any of Q7_1–Q7_4, QID11_1–QID11_6, or QID12_1–QID12_8. An all-volunteer network has no staff, so the answers cannot both be true.
- `Completed in under 2 minutes`: Qualtrics estimates the survey takes about 9 minutes. Under 2 minutes (120 seconds) is too fast to have read the questions. Responses between 2 and 5 minutes are not excluded but are flagged for review in Step 5.

**P — Probable** (requires human review of text)
Applied to all non-E rows where at least one open-text field contains a response, unless that text repeats another respondent's wording (see Q below). Filling in a free-response field is a strong indicator of human engagement. However, the text content should still be reviewed for AI-generated language.

The open-text fields checked are: Q1_7_TEXT (other role), Q2_4_TEXT (other org type), Q5_4_TEXT (other work type), Q6_4_TEXT (other work focus), Q8_9_TEXT (other opposition source), Q9_8_TEXT (other opposition form), Q12_8_TEXT (other opportunity), Q13_8_TEXT (other tension), Q15_7_TEXT (other approach), Q16_7_TEXT (other strategy), Q17a_7_TEXT (other shift), Q19_7_TEXT (other framework change), Q20 (collaboration description), Q21a (cross-sector collaboration description), Q22 (open narrative response).

**Q — Questionable**
All non-E rows that meet either of these conditions:

- `No open-text responses`: The respondent completed only multiple-choice items, which automated responses can replicate easily.
- `Repeated text`: At least one open-text answer shares a run of 4 or more identical consecutive words with another respondent's answer to the same question. Matching ignores capitalization and punctuation. Copy-pasted or bot-generated submissions often reuse the same phrases, while independent human answers rarely match word-for-word for that long. Answers shorter than 4 words are never flagged by this rule.

See Step 5 for additional detail flags that help prioritize which Q-flagged responses to review first.

**Reason columns**
So you can see why each row got its flag, the output includes one 0/1 column per rule (1 = the rule applies) plus a readable summary column:

| Column | Meaning |
|---|---|
| `E_survey_preview` | Status is "Survey Preview" |
| `E_q1_blank` | Q1 is blank |
| `E_under_18_answered` | Fewer than 18 of 27 questions answered |
| `E_volunteer_staff_conflict` | All-volunteer network in Q1 but "Staff" checked |
| `E_under_2_min` | Completed in under 2 minutes |
| `Q_no_text` | No open-text responses |
| `Q_repeated_text` | Open-text answer repeats another respondent's wording |
| `repeated_text_cols` | Which open-text columns had the repeated wording |
| `questions_answered` | Number of the 27 questions answered |
| `flag_reason` | Every reason that applies, separated by semicolons |

The reason columns list every rule a row breaks, even though the row gets only one flag. For example, a row can have both `E_q1_blank = 1` and `E_under_2_min = 1`.

```{r define-question-groups}
# Each entry lists all columns belonging to one survey question.
# A question counts as "answered" if any of its columns contains a response.
question_groups <- list(
  Q1     = "Q1",
  Q2     = "Q2",
  Q3     = "Q3",
  Q23_states = "Q23",
  Q5     = "Q5",
  Q6     = "Q6",
  Q7     = c("Q7_1","Q7_2","Q7_3","Q7_4"),
  QID11  = c("QID11_1","QID11_2","QID11_3","QID11_4","QID11_5","QID11_6"),
  QID12  = c("QID12_1","QID12_2","QID12_3","QID12_4","QID12_5","QID12_6",
             "QID12_7","QID12_8","QID12_9"),
  Q8     = "Q8",
  Q9     = "Q9",
  Q10    = c("Q10_1","Q10_2","Q10_3","Q10_4"),
  Q12    = "Q12",
  Q13    = "Q13",
  Q14    = c("Q14_1","Q14_2","Q14_3","Q14_4"),
  Q15    = "Q15",
  Q15a   = c("Q15a_1","Q15a_2","Q15a_3","Q15a_4","Q15a_5","Q15a_6","Q15a_7"),
  Q16    = "Q16",
  Q16a   = c("Q16a_1","Q16a_2","Q16a_3","Q16a_4","Q16a_5","Q16a_6","Q16a_7"),
  Q17    = "Q17",
  Q17a   = "Q17a",
  Q18    = "Q18",
  Q19    = "Q19",
  Q20    = "Q20",
  Q21    = "Q21",
  Q21a   = "Q21a",
  Q22    = "Q22"
)
```

```{r assign-flags}
text_cols <- c(
  "Q1_7_TEXT", "Q2_4_TEXT",  "Q5_4_TEXT",   "Q6_4_TEXT",
  "Q8_9_TEXT", "Q9_8_TEXT",  "Q12_8_TEXT",  "Q13_8_TEXT",
  "Q15_7_TEXT","Q16_7_TEXT", "Q17a_7_TEXT", "Q19_7_TEXT",
  "Q20", "Q21a", "Q22"
)

questions_answered <- apply(df, 1, function(r) {
  sum(sapply(question_groups, function(cols) any(nchar(trimws(r[cols])) > 0)))
})

has_text <- apply(df[, text_cols], 1, function(r) any(nchar(trimws(r)) > 0))

# Contradiction: all-volunteer network in Q1, but "Staff" checked elsewhere
staff_cols <- c(paste0("Q7_", 1:4), paste0("QID11_", 1:6), paste0("QID12_", 1:8))
is_volunteer_network <- str_detect(df$Q1, fixed("Member of an all-volunteer network",
                                                 ignore_case = TRUE))
checked_staff <- apply(df[, staff_cols], 1, function(r) {
  any(str_detect(r, regex("\\bstaff\\b", ignore_case = TRUE)), na.rm = TRUE)
})
volunteer_staff_conflict <- is_volunteer_network & checked_staff

# Repeated text: 4+ identical consecutive words shared with another row's
# answer in the same column
word_runs <- function(x, n = 4) {
  if (is.na(x)) return(character(0))
  words <- str_split(str_squish(str_replace_all(tolower(x), "[^a-z0-9']+", " ")), " ")[[1]]
  words <- words[words != ""]
  if (length(words) < n) return(character(0))
  unique(sapply(seq_len(length(words) - n + 1),
                function(i) paste(words[i:(i + n - 1)], collapse = " ")))
}

repeated_by_col <- sapply(text_cols, function(col) {
  runs   <- lapply(df[[col]], word_runs)
  counts <- table(unlist(runs))          # each run counted once per row
  shared <- names(counts)[counts > 1]
  sapply(runs, function(r) any(r %in% shared))
})
repeated_by_col <- matrix(repeated_by_col, nrow = nrow(df),
                          dimnames = list(NULL, text_cols))
has_repeated_text <- rowSums(repeated_by_col) > 0

# One 0/1 column per rule, so the output shows why each row was flagged
df$questions_answered         <- questions_answered
df$E_survey_preview           <- as.integer(coalesce(df$Status == "Survey Preview", FALSE))
df$E_q1_blank                 <- as.integer(coalesce(trimws(df$Q1) == "", TRUE))
df$E_under_18_answered        <- as.integer(coalesce(questions_answered < 18, FALSE))
df$E_volunteer_staff_conflict <- as.integer(coalesce(volunteer_staff_conflict, FALSE))
df$E_under_2_min              <- as.integer(coalesce(df$duration_sec < 120, FALSE))
df$Q_no_text                  <- as.integer(!coalesce(has_text, FALSE))
df$Q_repeated_text            <- as.integer(has_repeated_text)
df$repeated_text_cols         <- apply(repeated_by_col, 1, function(r)
                                   paste(text_cols[r], collapse = "; "))

is_E <- with(df, E_survey_preview | E_q1_blank | E_under_18_answered |
                 E_volunteer_staff_conflict | E_under_2_min) == 1

df$response_flag <- case_when(
  is_E                    ~ "E",
  df$Q_repeated_text == 1 ~ "Q",
  df$Q_no_text == 0       ~ "P",
  TRUE                    ~ "Q"
)

reason_labels <- c(
  E_survey_preview           = "Survey preview",
  E_q1_blank                 = "Q1 blank",
  E_under_18_answered        = "Fewer than 18 questions answered",
  E_volunteer_staff_conflict = "All-volunteer network but checked staff",
  E_under_2_min              = "Completed in under 2 minutes",
  Q_no_text                  = "No open-text responses",
  Q_repeated_text            = "Repeated text"
)
df$flag_reason <- apply(df[, names(reason_labels)], 1, function(r)
  paste(reason_labels[as.integer(r) == 1], collapse = "; "))
```

```{r verify-flags}
cat("=== Response flag distribution ===\n")
print(table(df$response_flag, useNA = "ifany"))

cat("\n=== Rows meeting each rule (a row can meet more than one) ===\n")
for (col in names(reason_labels)) {
  cat(sprintf("  %-42s %d\n", reason_labels[[col]], sum(df[[col]])))
}
```

---

## Step 5: Suspicion Detail Flags

These three columns provide additional detail to help prioritize manual review of Q-flagged responses. They are calculated for all rows but are most relevant for Q entries. They do **not** change the E / P / Q assignment — they are supplementary information only.

**`flag_short_duration`** — Completion time under 5 minutes (300 seconds). Qualtrics estimates this survey takes about 9 minutes, so finishing in under 5 minutes, with 18 required questions and multiple open-ended items, suggests the respondent was not reading carefully. (Responses under 2 minutes are already marked E in Step 4; this flag is also set for them, but it matters most for the 2–5 minute range, which stays P or Q.)

**`flag_straightline_Q10`** — All four Q10 Likert scale items (measuring perceptions of right-wing threats) answered identically. Uniform responses across four conceptually distinct statements suggest inattentive or automated completion. Note: the raw data contains inconsistent capitalization in Q10 responses (e.g., "strongly disagree" vs. "Strongly agree"); `tolower()` normalizes this before comparison.

**`flag_straightline_Q14`** — Same straight-lining logic applied to Q14 (four Likert items measuring perceptions of internal organizational tensions).

**`suspicion_count`** — A simple sum of the three flags above, ranging from 0 to 3. Higher values indicate more reason to scrutinize the response during manual review.

```{r suspicion-flags}
# Flag 1: Completed in under 5 minutes (300 seconds)
df$flag_short_duration <- as.integer(df$duration_sec < 300)

# Flag 2: Straight-lining on Q10 (right-wing threat Likert scale)
q10_cols <- c("Q10_1","Q10_2","Q10_3","Q10_4")
df$flag_straightline_Q10 <- apply(df[, q10_cols], 1, function(r) {
  filled <- tolower(trimws(r[nchar(trimws(r)) > 0]))
  as.integer(length(filled) == 4 && length(unique(filled)) == 1)
})

# Flag 3: Straight-lining on Q14 (internal tensions Likert scale)
q14_cols <- c("Q14_1","Q14_2","Q14_3","Q14_4")
df$flag_straightline_Q14 <- apply(df[, q14_cols], 1, function(r) {
  filled <- tolower(trimws(r[nchar(trimws(r)) > 0]))
  as.integer(length(filled) == 4 && length(unique(filled)) == 1)
})

# Combined suspicion count
df$suspicion_count <- df$flag_short_duration +
                      df$flag_straightline_Q10 +
                      df$flag_straightline_Q14
```

```{r verify-suspicion}
cat("=== Suspicion flags among Q-flagged responses ===\n")
q_rows <- filter(df, response_flag == "Q")
cat(sprintf("  Short duration (<5 min):    %d\n", sum(q_rows$flag_short_duration,   na.rm = TRUE)))
cat(sprintf("  Straight-line Q10:          %d\n", sum(q_rows$flag_straightline_Q10, na.rm = TRUE)))
cat(sprintf("  Straight-line Q14:          %d\n", sum(q_rows$flag_straightline_Q14, na.rm = TRUE)))
cat(sprintf("  2 or more flags raised:     %d\n", sum(q_rows$suspicion_count >= 2,  na.rm = TRUE)))
```

---

## Step 6: Final Summary

```{r final-summary}
cat("=== Final dataset dimensions ===\n")
cat(sprintf("  Rows:    %d\n", nrow(df)))
cat(sprintf("  Columns: %d\n", ncol(df)))

cat("\n=== Response flag counts ===\n")
flag_tbl <- table(df$response_flag, useNA = "ifany")
print(flag_tbl)
cat(sprintf("\n  Total accounted for: %d\n", sum(flag_tbl)))
```

---

## Step 7: Save Output

The cleaned and flagged dataset is saved as a new CSV in the same Downloads folder as the raw file. The original file is not modified.

```{r save-output}
output_path <- file.path(dirname(file_path), "labor_repro_survey_flagged.csv")
write.csv(df, output_path, row.names = FALSE)
cat(sprintf("Saved to: %s\n", output_path))
cat(sprintf("Dimensions: %d rows x %d columns\n", nrow(df), ncol(df)))
```
